Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-15 manuscript.tex
Update on Overleaf.

\change[Antoine]{To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.}{Even if technology lags far behind biology, this introduction demonstrates the importance of timing in neural processes and we will further review the role of precise spike timing in neural assemblies. We start by reviewing the different hypotheses that aim at deciphering the neural code with spatio-temporal spiking motifs. After listing some biological evidence for the use of precise spike timing we review some computational models and neuromorphic technics that add this temporal dimension to their computations.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons~\citep{bruno_cortex_2006}. \noet{neuron-centric view from the reviewer comment, make some changes to broaden the proposed vision. Also change the very specific example given after, or put in context (works by Sophie Dénève or Ballard and Jehee)}

>>>>

\change[Antoine]{To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.}{Even if technology lags far behind biology, this introduction demonstrates the importance of timing in neural processes and we will further review the role of precise spike timing in neural assemblies. We start by reviewing the different hypotheses that aim at deciphering the neural code with spatio-temporal spiking motifs. After listing some biological evidence for the use of precise spike timing we review some computational models and neuromorphic technics that add this temporal dimension to their computations.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\add[Antoine]{In this first section, we have introduced the notion of rate coding and demonstrated that spike timing can also carry information. There are multiple experimental evidences that neurons use different strategies to communicate.}~\cite{Maunsell_Functional_1983} \add[Antoine]{demonstrate that neurons of visual area MT in macaque monkeys encode motion direction through population activity, offering robustness to single neuron activity fluctuations. For rapid image processing, it has been observed that neurons in the salamander retina represent the spatial structure of an image with the relative timing of their first spikes}~\citep{gollisch_rapid_2008}\add[Antoine]{. In the primary visual area of macaque monkeys, the phase of firing convey information about naturalistic stimuli that cannot be described by firing rates alone}~\citep{montemurro_phase-of-firing_2008}\add[Antoine]{. Among a variety of strategies to communicate, so far we have also evidence that neurons can use correlation coding}~\citep{decharms_primary_1996}\add[Antoine]{ or sparse coding}~\citep{vinje_sparse_2000}. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intracortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons~\citep{bruno_cortex_2006}. \noet{neuron-centric view from the reviewer comment, make some changes to broaden the proposed vision. Also change the very specific example given after, or put in context (works by Sophie Dénève or Ballard and Jehee)}
